\\Large The heat is on: Rising temperatures alter how and where mammals move

### 2025-09-03 writing/manuscript.Rmd
reviewed results

\begin{equation}\label{eq:hrsf}
\begin{cases}
D \sim \text{Pois}(\lambda) \\
\mathbb E(D) = \mathbb V(D) = \lambda \\
\begin{aligned}
\text{log}(\lambda) = & f_{1}(\texttt{forest\_perc}) + f_{2}(\texttt{elevation\_m}) + f_{3}(\texttt{dist\_water\_m}) + \\
& Z_\texttt a + f_{4, \texttt{a}}(\texttt{forest\_perc}) + f_{5, \texttt{a}}(\texttt{elevation\_m}) + f_{6, \texttt{a}}(\texttt{dist\_water\_m}) + \\
& f_{7}(\texttt{forest\_perc}, \texttt{temp\_c}) + f_{8}(\texttt{elevation\_m}, \texttt{temp\_c}) + \\
& f_{9}(\texttt{dist\_water\_m}, \texttt{temp\_c}) + f_{10}(\texttt{temp\_c}) + f_{11,\texttt a}(\texttt{temp\_c}))
\end{aligned}
\end{cases},
\end{equation}

\noindent where the species-level indices are omitted for readability, but each term in the model can be assumed to be species-specific. Smooth effects of percent forest cover (\texttt{forest\_perc}), elevation (\texttt{elev\_m}), and distance to water (\texttt{dist\_water\_m}) accounted for the species-level selection strength for each resource. A Gaussian random effect for each individual animal ($Z_\texttt a$) corrected for uneven sampling across individuals, while factor-smooth interaction terms for each animal ($f_{j,\texttt a}$) accounted for animal-level resource selection [i.e., individual-level deviations from the species-level estimate$;$ @jeltsch_need_2025]. Tensor product interaction terms of the three resources and temperature (\texttt{temp\_c}) estimated the smooth change in resource selection at different temperatures. Finally, marginal smooth terms of temperature and factor-smooth interaction terms of temperature and animal accounted for species- and individual-level sampling biases at different temperatures (e.g., sampling more during warm periods).

@@

Near 0$^\circ$C, wolves selected for dense forest cover ($\gtrapprox$ 50%), elevations near 1 km, and distances from water < 10 km; elk selected for forest cover near 50%, elevations between 1 and 2 km, and distances from water of 5-15 km; mountain goats selected for sparse (< 50%) forest cover, elevations between 1 and 2 km; cougars selected for dense forest cover, an elevation of ~ 1 km, and distances from water < 10 km; boreal caribou selected for intermediate to dense forest cover, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water < 5 km; and grizzly bears selected for sparse forest cover (25-50%), elevation near 1 km, and distances from water < 2 km.

There was relatively strong agreement between models with and without temperature (Figs. B3, and B14), but including temperature always resulted in better fits. All analyses of deviance showed that temperature significantly affected movement behaviour (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Table B2 and following pages in Appendix B). Accounting for temperature resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.

@@

\noindent The effects of temperature on movement rates varied in both direction and magnitude across species, even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs. B4-B6). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models had good in-sample prediction (Fig. B7) and explained reasonably high proportions of the deviance (79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch. 11 of @mcelreath_statistical_2020]. All species adapted their daily and seasonal movement behaviour to changes in temperature (Fig. \@ref(fig:dist)). The clearest example of this was cougars. In mid summer, they moved mostly in the evening if temperatures were cool and mostly in the early morning if temperatures were hot. Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" depended on the time of year. However, uncertainties around the estimated effects were often higher at extreme temperatures due to data scarcity. All species' estimated probability of moving decreased with sampling intervals approximately above 1 hour, and most species' estimated speed when moving decreased monotonically with sampling interval, although the estimated trends were highly uncertain for some species (Fig. B8).

```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the estimated effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}). Predictions in the surface plots extend to 10% of the range away from each datum. The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to $2^{\\pm 2}$ for ease of readability."}
# not specifying dt, or doy for A and time for B because change is relative
knitr::include_graphics('../figures/distance.png')
```

@@

Species' relative selection strength (RSS) was generally strongest for elevation and weakest for forest cover, but species' RSS for resources depended on temperature (Fig. \@ref(fig:hrsfs)). Changes in RSS with temperature were also strongest for elevation and generally weakest distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of cougars, exhibited a clear temperature-dependent shift in their preference for forest cover. At higher temperatures, wolves, mountain goats, and grizzly bears became less selective for forest cover, while elk and caribou shifted towards more intermediate forest cover without much of a change in preference width. All species shifted elevationally with temperature, although boreal caribou's selection strength for elevation was less sensitive to elevation. Overall, as temperatures rose, elk, mountain goats, and cougars increased in elevation, while boreal wolves, southern mountain caribou, and grizzly bears decreased in elevation. Most species generally remained within 5 km of water, and temperature did not affect their selection strength as much as for the other two resources. Again, estimated RSS values were generally most uncertain at extreme temperatures (Fig. B15).

```{r hrsfs, fig.cap="\\textbf{Temeperature strongly affects mammals' habitat selection.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature. The grey contours indicate the extent of each species' observed data. RSS values were re-centered and capped to $2^{\\pm 2}$ to improve readability."}
knitr::include_graphics('../figures/hrsf-surface-plots.png')
```

\noindent The predicted changes in distance traveled varied across species in both direction and magnitude, but worse climate-change SSPs always corresponded to greater absolute changes (Fig. \@ref(fig:t-dist)). Additionally, species that were predicted to move less often did not necessarily have lower speeds when moving, and vice versa (Figs. B9 and B10). Overall, absolute changes by 2100 under the best-case SSP were small (approximately 0% to +4%), while under the worst-case SSP absolute changes ranged from ~2% (grizzly bears) to ~24% (cougars). Notably, while the models estimated that grizzly bears would move substantially less (if at all) in winter (Figs. \@ref(fig:dist), B4-B6), the models did not explicitly account for changes in hibernation phenology. Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology. Projected changes also varied spatially due to heterogeneity in projected climate change (Fig. \@ref(fig:s-dist)). Again, absolute changes were generally greatest under worse SSPs, but the direction of change at each location also varied across SSPs (most visible in cougars).

@@

```{r t-dist, fig.cap="\\textbf{The direction and magnitude of changes in movement rates due to climate change varies among species, but worse climate-change scenarios result in the greatest change.} Lines indicate the median projected change in distance travelled (probability of moving times speed when moving) due to changes in temperature within each species' observed extent. Shaded areas indicate the range of predicted values between the $95^\\text{th}$ and $5^\\text{th}$ percentiles. Changes are relative to the predicted mean distances travelled in 2025 across the four Shared Socioeconomic Pathways (SSPs). If the intervals are fully above the dashed line, at least 90% of the estimated means increased, relative to the each pixel's average of the four climate SSPs in 2025. Similar conclusions can be drawn regarding a decrease if the ribbons are below the dashed line. The projections only account for changes in movement behaviour (i.e., movement frequency and speed) and ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently.} The color scale indicates the predicted changes in distance traveled in 2100, relative to each pixel's average distance in 2025 across all four scenarios (i.e., not relative to other pixels). Values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.50 to 1.56). The predictions only account for the predicted temperature throughout the province and ignore environmental factors such as terrain slope, soil type, and forest density. All maps are shown with a BC Albers projection (EPSG:3005)."}
if(FALSE) range(readRDS('../data/cc-hgam-projections-local-2100.rds')$d)
knitr::include_graphics('../figures/local-distance-2100.png')
```

Median RSS was projected to decrease over time within the each species' observed range, but changes were stronger under worse SSPs (Fig. \@ref(fig:s-hrsfs)). Decreases were most pronounced in the bottom fifth percentile and most severe for elk, southern mountain caribou, and mountain goats. Of all the species, only cougars showed a clear increase in RSS for areas within their current range. The change in RSS between 2025 and 2100 varied spatially for all species (Fig. \@ref(fig:s-hrsfs)). All species were predicted to exhibit a decrease in RSS throughout most of their observed habitats, although cougars were predicted to increase their selection for patches with higher altitude.

@@

```{r t-hrsfs, fig.cap="\\textbf{All species are expected to exhibit a decrease in median relative selection strength (RSS) for their current range, irrespective of climate change scenario, but decreases are stronger in the worse-case scenario.} Lines indicate the projected change in median RSS within each species' observed extent within BC, while the ribbons indicate the range between the $95^\\text{th}$ and $5^\\text{th}$ percentiles in RSS. Changes are relative to each location's mean RSS in 2025 across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions.png')
```

```{r s-hrsfs, fig.cap="\\textbf{Climate change will impact each species' habitat relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios (i.e., not relative to other pixels). Values < 1 indicate a decrease and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.67 and 1.5 ($\\approx 2^{\\pm 0.585}$; original data ranged 0.66 to 1.86). All maps are shown with a BC Albers projection (EPSG:3005)."}
if(FALSE) range(readRDS('../data/cc-hrsf-projections-local-2100.rds')$l)
knitr::include_graphics('../figures/local-rss-2100.png')
```

@@

@@

The diversity in predicted changes in movement behaviour by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig. \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. For example, the width of the prediction intervals in Fig. \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas [@jeltsch_need_2025]. Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025]. Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].

>>>>

\begin{equation}\label{eq:hrsf}
\begin{cases}
C \sim \text{Pois}(\lambda) \\
\mathbb E(C) = \mathbb V(C) = \lambda \\
\begin{aligned}
\text{log}(\lambda) = & f_{1}(\texttt{forest\_perc}) + f_{2}(\texttt{elevation\_m}) + f_{3}(\texttt{dist\_water\_m}) + \\
& Z_\texttt a + f_{4, \texttt{a}}(\texttt{forest\_perc}) + f_{5, \texttt{a}}(\texttt{elevation\_m}) + f_{6, \texttt{a}}(\texttt{dist\_water\_m}) + \\
& f_{7}(\texttt{forest\_perc}, \texttt{temp\_c}) + f_{8}(\texttt{elevation\_m}, \texttt{temp\_c}) + \\
& f_{9}(\texttt{dist\_water\_m}, \texttt{temp\_c}) + f_{10}(\texttt{temp\_c}) + f_{11,\texttt a}(\texttt{temp\_c}))
\end{aligned}
\end{cases},
\end{equation}

\noindent where $O$ indicates whether an animal was observed ($O = 1$) or not ($O = 0$), and the species-level indices are omitted for readability, but each term in the model can be assumed to be species-specific. Smooth effects of percent forest cover (\texttt{forest\_perc}), elevation (\texttt{elev\_m}), and distance to water (\texttt{dist\_water\_m}) accounted for the species-level selection strength for each resource. A Gaussian random effect for each individual animal ($Z_\texttt a$) corrected for uneven sampling across individuals, while factor-smooth interaction terms for each animal ($f_{j,\texttt a}$) accounted for animal-level resource selection [i.e., individual-level deviations from the species-level estimate$;$ @jeltsch_need_2025]. Tensor product interaction terms of the three resources and temperature (\texttt{temp\_c}) estimated the smooth change in resource selection at different temperatures. Finally, marginal smooth terms of temperature and factor-smooth interaction terms of temperature and animal accounted for species- and individual-level sampling biases at different temperatures (e.g., sampling more during warm periods).

@@

Across all species, Relative Selection Strength (RSS) was weakest for forest cover and strongest for elevation. At temperatures near 0$^\circ$C, wolves selected for forest cover ($\gtrapprox$ 50%), elevations near 1 km, and distances from water < 5 km; elk selected for intermediate forest cover ($\approx$ 50%), elevations between 1 and 2 km, and distances from water of 10-15 km; mountain goats selected for sparse forest cover (< 25%), elevations near 1.5 km, and distances from water < 5 km; cougars selected for dense forest cover (> 75%), an elevation of ~ 1 km, and distances from water < 7.5 km; boreal caribou selected for forest cover between 50% and 75%, elevations near 500 m, and distances from water < 10 km; southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water $\lessapprox$ 5 km; and grizzly bears selected for relatively sparse forest cover (25-50%), elevation between 1 and 2 km, and distances from water < 3 km.

There was relatively strong agreement between models with and without temperature (Figs. B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Table B2 and subsequent pages in Appendix B). Accounting for temperature even resulted in a simpler HRSF for mountain goats (lower model degrees of freedom) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.

@@

\noindent Species' changes in movement rates to temperature varied in both direction and magnitude (Figs. \@ref(fig:dist), B4-B6), even after accounting for differences in daily and seasonal activity (e.g., sleeping, migration, hibernation; Figs. B4-B6). Smooth interaction terms were well-behaved and indicated clear shifts in activity over time of day and day of year for all species. The models had good in-sample prediction (Fig. B7) and explained reasonably high proportions of the deviance [79.3% for the gamma model and 10.7% for the binomial model, which is relatively high for a binomial model with binary responses: see ch. 11 of @mcelreath_statistical_2020]. All species adapted their daily and seasonal movement behaviour to changes in temperature (Fig. \@ref(fig:dist)). The clearest example of this was cougars. In late spring (June 1^st^), they moved from evening to early morning if temperatures were below $20^\circ$C, but if temperatures were above 20$^\circ$C they moved mostly between 3:00 and 6:00 AM. Throughout the year, they tended to move more when it was colder, but what they perceived as "colder" was relative to the time of year. All species' estimated probability of moving and speed when moving decreased with sampling intervals above 1 hour, except for cougars' speed, although the estimated trends were highly uncertain (Fig. B8). Uncertainties around the estimated effects were generally higher at extreme temperatures due to lower data availability (Figs. B4A, B5A, and B6A).

```{r dist, fig.cap="\\textbf{Temperature is a strong determinant of how far and when mammals travel.} The fill indicates the effects of temperature on the relative change in distance travelled (probability of moving times mean speed when moving) over time of day on June $1^{\\text{st}}$ (\\textbf{A}) and day of year at 12:00 PDT (\\textbf{B}). Predictions in the surface plots extend to 10% of the range away from each datum. The color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to $2^{\\pm 2}$ for ease of readability."}
# not specifying dt and doy for A and time for B because change is relative
knitr::include_graphics('../figures/distance.png')
```

@@

Species' RSS was generally strongest for elevation and weakest for forest cover, but RSS depended significantly on temperature for all species (all p-values $< 2.2\times10^{-16}$; Fig. \@ref(fig:hrsfs)). Changes in RSS with temperature were strongest for elevation and generally weakest for distance from water, but there were no common trends across all species for any of the three resources. All species, with the exception of cougars, exhibited clear temperature-dependent shifts in their preference for forest cover. At higher temperatures, wolves relaxed their preference for forested areas, while mountain goats relaxed their preference open areas (cover < 50%). As temperatures warmed, elk and boreal caribou shifted towards more forest cover closer to 50%, while southern mountain caribou and grizzly bears selected for areas with 50% forest cover or less. All species shifted elevationally with temperature, although boreal caribou ha a much sharper change in RSS across elevation. Elk, mountain goats, and cougars increased in elevation with temperature, while wolves, southern mountain caribou, and grizzly bears decreased in elevation. Wolves, elk, and southern mountain caribou moved closer to water with temperature, while mountain goats, cougars, and grizzly bears moved somewhat further away from water but remained mainly within 5-10 of water. As with movement rates, estimated RSS was generally most uncertain at extreme temperatures, for which data was scarcer (Fig. B15).

```{r hrsfs, fig.cap="\\textbf{Mammals' habitat selection depends on ambient temperature.} Estimated relative selection strength (RSS) for forest cover (\\%), elevation (km), and distance from water (km) as a function of temperature. The grey contours indicate the extent of each species' observed data. RSS values were re-centered and capped to $2^{\\pm 2}$ to improve readability."}
knitr::include_graphics('../figures/hrsf-surface-plots.png')
```

## Predicted changes in movement behaviour during the 21^st^ century

\noindent Predicted changes in movement rates with future climate change varied across species in both magnitude and direction, but worse SSPs always corresponded to greater absolute changes (Fig. \@ref(fig:t-dist)). Additionally, species that were predicted to move less often did not necessarily have lower speeds when moving, and vice versa (Figs. B9 and B10). Absolute relative changes in distance travelled were small under the best-case SSP (0-4% change in 2100 relative to 2025). Under the worst-case SSP, absolute changes by 2100 (relative to 2025) ranged from ~2% (grizzly bears) to ~24% (cougars). Notably, while the models estimated that grizzly bears would move substantially less (if at all) in late fall and winter (Figs. \@ref(fig:dist), B4), the models did not explicitly account for changes in hibernation phenology. Consequently, the climate change projections do not explicitly account for changes in energetic needs or physiology. Projected changes in 2100 varied spatially due to spatial heterogeneity in climate change projections (Fig. \@ref(fig:s-dist)). Again, absolute changes were generally greatest under worse SSPs, but the direction of change at each location also varied across SSPs (most visible in cougars).

@@

```{r t-dist, fig.cap="\\textbf{Species will adapt their movement rates differently in response to climate change, but worse climate-change scenarios will result in the greatest change.} Bold lines indicate the median change in distance travelled (probability of moving times speed when moving) due to predicted changes in temperature within each species' observed extent. Shaded areas indicate the $95^\\text{th}$ and $5^\\text{th}$ percentiles. Changes are relative to the mean predicted distance travelled at each location in 2025, averaged across the four Shared Socioeconomic Pathways (SSPs). Values > 1 indicate an increase, while values < 1 indicate a decrease. The projections only account for changes in movement frequency and speed, and they ignore changes in physiology or movement costs."}
knitr::include_graphics('../figures/distance-travelled-local-cc-predictions.png')
```

```{r s-dist, fig.cap="\\textbf{Climate change will impact each species' movement rates differently, but changes will also vary spatially.} The color of each pixel indicates the predicted changes in distance traveled in 2100 at that pixel, relative to the value in 2025, averaged across all four scenarios. Values < 1 indicate a decrease, and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling). The predictions only account for the predicted temperature throughout the areas and ignore environmental factors such as terrain slope, soil type, and forest density. All maps are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
if(FALSE) range(readRDS('../data/cc-hgam-projections-local-2100.rds')$d)
knitr::include_graphics('../figures/local-distance-2100.png')
```

Median RSS was projected to decrease over time within the each species' observed range, but, again, changes were stronger under worse SSPs (Fig. \@ref(fig:t-rss-cc)). Decreases were most pronounced in areas with the lowest RSS and most severe for elk, mountain goats, pumas, and southern mountain caribou. Changes for boreal wolves and caribou were relatively negligible. Elk, cougars, and grizzly bears were predicted to increase their selection strength for top-RSS areas, and elk, mountain goats, cougars, and southern mountain caribou were predicted to further decrease their selection for areas with low RSS. Unsurprisingly, the change in RSS between 2025 and 2100 also varied spatially for all species (Fig. \@ref(fig:s-rss-cc)). Overall, RSS decreased throughout most of each species' current range, although elk, cougars, and bears were predicted to increase their selection for higher-altitude habitats. Still, none of the species were projected to increase RSS throughout their habitat (\@ref(fig:density-hrsfs)).

@@

```{r t-rss-cc, fig.cap="\\textbf{Species will adapt their habitat selection differently in response to climate change, but worse climate-change scenarios will result in the greatest change.} Bold lines indicate the change in median RSS due to predicted changes in temperature within each species' observed extent. Shaded areas indicate the $95^\\text{th}$ and $5^\\text{th}$ percentiles in change in RSS. Values > 1 indicate an increase, while values < 1 indicate a decrease. Changes are relative to each location's mean RSS in 2025, averaged across the four Shared Socioeconomic Pathways (SSPs)."}
knitr::include_graphics('../figures/rss-local-cc-predictions.png')
```

```{r s-rss-cc, fig.cap="\\textbf{Climate change will impact each species' relative selection strength (RSS) differently.} The color scale indicates the predicted changes in RSS in 2100, relative to each location's average RSS in 2025 across all four scenarios (i.e., not relative to other pixels). Values < 1 indicate a decrease and values > 1 indicate an increase. For ease of readability, the color bar is on the $\\log_2$ scale to help visualize patterns in doubling, and values are capped to 0.8 and 1.25 ($\\approx 2^{\\pm 0.322}$; original data ranged 0.71 to 1.93). All maps are shown in the BC Albers Equal Area Conic projection (EPSG:3005)."}
if(FALSE) range(readRDS('../data/cc-hrsf-projections-local-2100.rds')$l)
knitr::include_graphics('../figures/local-rss-2100.png')
```

```{r density-hrsfs, fig.cap="\\textbf{Climate change will cause species to decrease their selection strength for their current habitats, but worse climate change scenarios will have the greatest impact.} The density plots indicate each species' change in RSS in 2100 for their current habitat (Fig. 8), relative to each location's RSS in 2025. Values < 1 indicate a decrease and values > 1 indicate an increase. For ease of readability, values > 1.3 were changed to 1.3 (original maximum: 1.93)."}
knitr::include_graphics('../figures/local-rss-2100-density.png')
```

@@

***Overall, estimated selection strength was weakest for forest cover and strongest for elevation. Elevation is the only raster that doesn't change noticeably over time, so time-varying rasters of the other two would have likely given stronger (and better) estimates.***

@@

The diversity in predicted changes in movement behaviour by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig. \@ref(fig:s-rss-cc) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. For example, the width of the prediction intervals in Fig. \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas [@jeltsch_need_2025]. Population-level means that ignore such variation can greatly limit the efficacy of conservation projects [@muff_accounting_2020; @mortlock_sleep_2024; @silva_too_2025]. Additionally, accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations undergo a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].
